---
title: "MortgageMarketAnalysis"
output:
  html_document:
    df_print: paged
date: '2022-06-16'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this analysis is to compare relative performance of mortgage
bonds to US gov't bonds in order to see how the FOMC rate hike on
6/15/2022 and the abysmal housing report on 6/16/2022 are affecting the
mortgage markets. In this analysis, I will import the data (Excel sheet
attached), perform some basic exploratory data analysis, and then from
the data analysis, decide on a feasible statistical method to extract
insight into the current mortgage market.

First we begin by loading in the necessary data and libraries.
Additionally, I filter out unneeded columns that are blank and limit the
data to returns from 2022. An important thing to note is that I
performed a log difference approximation in order to be able to compare
the returns across different bonds.

```{r message=FALSE, warning=FALSE}
# Clear the Workspace
rm(list=ls())

library(tidyverse)
library(readxl)
library(devtools)

rate_mbs_selloff <- read_excel("rate mbs selloff.xlsx")

#filter out NA columns by position
data1 <- rate_mbs_selloff %>% select(
  1,2,5,8,11,14,17
)

#only want 2022 data
data1 <- data1 %>% slice(1:116)#116

#check to make sure it works
View(data1)
```

After this is done, we can perform exploratory data analysis and
plotting of our data. The main focus of our EDA will be to explore the
correlation of the yields.

```{r}
#start EDA by plotting all line with ggplot2
library(ggplot2)

plot1 <- ggplot()+
  geom_line(data = data1,
            aes(x = Date, y = us_govt_2y),
            color='blueviolet')+
         
  geom_line(data = data1,
            aes(x = Date, y = us_govt_5y),
            color='pink')+
            
  geom_line(data = data1,
            aes(x = Date, y = us_govt_10y),
            color='cyan2')+
            
  geom_line(data = data1,
            aes(x = Date, y = fannie_4cpn),
            color='blue')+

  geom_line(data = data1,
            aes(x = Date, y = Mid_Yield_To_Convention),
            color='green')+
  
  geom_line(data = data1,
            aes(x = Date, y = fannie_5cpn),
            color='red')+
  
  theme_classic()+
  labs(title = 'Comparison of Bonds in 2022')+
  xlab('Date')+
  ylab('Bond Price')

plot1
```

Looking at the plot above, it appears that the bonds might be
correlated. In this case, principal component analysis will be a helpful
statistical method to use, as it is a dimension-reduction tool that is
used to reduce a large set of possibly correlated variables into a
smaller set number of uncorrelated variables(called the principal
components). These principal components will explain most of the
variation of the large set of variables. The first principle component
explains the maximum percentage of the total variance of interest
change. The second principal component is designed to be linearly
independent of the first and explains the maximum percentage of the
remaining variance, and so on and so forth.

In order to conduct analysis on the data, we need to check to see if it
is stationary, and if it is, we need to difference it.

```{r}
#remove dates to look at data
data4 <- data1 %>% select(
  2,3,4,5,6,7
)

#we can run acf and partial acf test on the 2 year Treasury Note and the Fannie 4% coupon to see if the data is non stationary
par(mfrow=c(1,2))
acf(data4$us_govt_2y)
pacf(data4$us_govt_2y)

par(mfrow=c(1,2))
acf(data4$fannie_4cpn)
pacf(data4$fannie_4cpn)
```

Looking at these results, the time-series for these two bonds and by
extent the rest of the bonds, are not stationary as the ACF decreases
gradually. We can now perform an Augmented Dickey Fuller (ADF) test to
identify the non-stationarity of the time-series. If the p-value from
the test is less than some significance level, then we will fail to
reject the null hypothesis and conclude that the time-series is
non-stationary.

```{r message=FALSE, warning=FALSE}
library(tseries)
adf.test(data4$fannie_4cpn)
```

The p-value comes out to 0.9484, meaning that we fail to reject the null
hypothesis and conclude that the data is non-stationary. As the
time-series shows a possible drift, let us check.

```{r}
library(urca)

test1 <- ur.df(data4$fannie_4cpn, type='drift',lags = 4)
summary(test1)
```

Upon running this Dickey-Fuller unit root with drift test, it can be
seen that interpreting the test is time consuming to determine if there
is drift involved in the data. Therefore, I used a function that can be
found at
<https://gist.github.com/hankroark/968fc28b767f1e43b5a33b151b771bf9> to
interpret the output.

```{r}
source("~/interp_urdf.R")

interp_urdf(test1, level = "5pct")
```

As can be seen, this shows that the Fannie 4% coupon bond contains no
drift.

Based on the analysis above, principle component analysis appears to be
an useful model. This allows us to reduce the dimensionality of the
different bonds and then analyze how the bond prices are functioning
related to each other. The principal components generated by the model
will be able to explain the vast majority of the yield curve while
remaining uncorrelated to each other.

The graph above shows that the bonds shown above appear to be highly
correlated. This is a problem if we would like to perform principal
component analysis as PCA requires a stationary time-series, which means
that a invariant of the market must be produced. According to
Meucci(2005), taking the first difference will be sufficient. Therefore,
we can establish that Rt,k = yield i,k - yield i-1,k as the yield return
of maturity k at time t.

```{r}

#remove dates to difference the series
data3 <- data1 %>% select(
  2,3,4,5,6,7
)
data3 <- data.frame(data3)

#difference the series once
returns <- as.data.frame(lapply(data3, diff, lag=1))
returns <- data.frame(returns)

#remove last row of data1 as we have a differenced time-series now
data1 <- data1 %>% slice(1:115)#115

#add the date column back to the dataset
returns$Date <- data1$Date


#plot the data to check that it worked
plot2 <- ggplot()+
  geom_line(data = returns,
            aes(x = Date, y = us_govt_2y),
            color='blueviolet')+
         
  geom_line(data = returns,
            aes(x = Date, y = us_govt_5y),
            color='pink')+
            
  geom_line(data = returns,
            aes(x = Date, y = us_govt_10y),
            color='cyan2')+
            
  geom_line(data = returns,
            aes(x = Date, y = fannie_4cpn),
            color='blue')+

  geom_line(data = returns,
            aes(x = Date, y = Mid_Yield_To_Convention),
            color='green')+
  
  geom_line(data = returns,
            aes(x = Date, y = fannie_5cpn),
            color='red')+
  
  theme_classic()+
  labs(title = 'Comparison of Bonds in 2022 (Differenced Time-Series)')+
  xlab('Date')+
  ylab('Bond Difference')

plot2
```

We need to confirm the data is now stationary with an ADF test.

```{r message=FALSE, warning=FALSE}
adf.test(returns$fannie_4cpn)
```

The p-value = 0.01, meaning we reject the null hypothesis that the data
is non-stationary.

Next, we can check the yield return matrix for any kind of correlation.
We expect to see a correlation in that if a bond with two year maturity
starts to pay less, there should be an impact on a bond with 5 year
maturity. The larger the difference in maturities, the lower the
correlation should be. Additionally, it is good to look at the matrix to
make sure there is some correlation, else PCA will not help reduce
dimensionality.

```{r}
#remove dates to look at correlation
cor.returns <- returns %>% select(
  1,2,3,4,5,6
)

options(digits = 2) #better visualization
cor(cor.returns) #correlation matrix of yield returns
```

As can be seen by a quick glance, this matrix is highly correlated. This
confirms that Treasury notes and mortgage rates are highly correlated
and will generally move in tandem with each other. This is an
economically sound principle, as it has been observed that interest
rates tend to become more synchronized in a distressed market
environment, as the current market is slowly becoming.

```{r}
model <- prcomp(cor.returns, scale = TRUE, center = TRUE)
summary(model)
# first principal component sign flip so it's an upward shift in yield
model$x[,1] <- -model$x[,1]
# also have to flip sign for eigenvector
model$rotation[,1] <- -model$rotation[,1]
 
# how much variance is captured in first n eigenvectors?
variance1 <- cumsum(model$sdev^2)/sum(model$sdev^2)
print(variance1)
```

```{r}
par(mfrow = c(1,2))
barplot(model$sdev^2, main = 'Eigenvalues of each component')
barplot(cumsum(model$sdev^2)/sum(model$sdev^2), main = 'Cumulative Explained Variance', ylab = 'Variance Explained')
```

As can be seen from the Cumulative Explained Variance graph, the
projection over three principal components can explain \~98% of the
variance across all 6 contracts. Next we can look at the factor loadings
to see how each one of them affect the returns of yields.

```{r}
par(mfrow = c(1,3))
hist(model$x[,1], breaks = 20, main = 'Distribution 1 component', xlab = paste('SD', round(model$sdev[1],2)))
hist(model$x[,2], breaks = 20, main = 'Distribution 2 component', xlab = paste('SD', round(model$sdev[2],2)))
hist(model$x[,3], breaks = 20, main = 'Distribution 3 component', xlab = paste('SD', round(model$sdev[3],2)))
```

```{r}

# plot loading factors
library(ggplot2)
ggplot(data.frame(cbind(term=c(1,2,3,5,7,10),model$rotation[,1:3]))) +
  geom_line(aes(x=term,y=PC1),colour="red") + geom_point(aes(x=term,y=PC1)) +
  geom_line(aes(x=term,y=PC2),colour="green") + geom_point(aes(x=term,y=PC2)) +
  geom_line(aes(x=term,y=PC3),colour="blue") + geom_point(aes(x=term,y=PC3)) +
  xlab("Maturity") +
  ylab("Shock Magnitude") +
  theme_bw()
```

This visualization captures the factor loadings. The red line is the
first principal component, and corresponds to a parallel move up and
down in the level of the entire yield curve. The green line is the
second principal component and is responsible for slope change and the
third component displays the convexity.

We can plot the first two principal components as a scatterplot to see
if there are any interesting features such as outliers or departures
from normality.

```{r}
library(ggfortify)
pca.plot <- autoplot(model, data = cor.returns, color = "Group")
pca.plot
```

There are a few outliers, but the first two principal components are
shown to behave within the limits of multivariate normality.

Now that it has been shown that the PCA is functioning correctly, we can
interpret our findings. In order to interpret each component, we can
compute the correlations between the original data and each principal
component in order to see which PCs are most correlated with the bonds.

```{r}
#cor.returns is the differenced dataset
#model$x is the PCA data

#take the correlation matrix to interpret the correlation between of the original variables and the PCs
comps.scaled <- pca.plot$data[,c(1:12)]
cor(comps.scaled[,3:8], comps.scaled[,c(1:2,9)])
```

This shows that the first principal analysis is positively correlated
with the bonds, meaning that the bonds vary together and as one goes
down, the others will decrease as well. This analysis allows us to
conclude that the bonds are highly correlated to each other and as PCA
analysis is shown to explain \~98% of the bond variance through three
principal components, we can use the residuals to determine how the
bonds are performing relative to each other.

The principal component analysis has served to confirm that mortgage
rates and Treasury bills are positively correlated and in times of
market distress, such as is being seen now, become even more positively
correlated. From this, it seems reasonable to expect that in the current
market, where Fed rate hikes are imminent and the recent housing market
report was abysmal, mortgage rates will continue to rise alongside
interest rates. In essence, the mortgage market currently hinges on
inflation. The expectation in the market right now is that the Fed will
keep increasing rates until it can be shown that inflation has peaked.
This means that Treasury bonds and by extension, mortgage rates, will
continue to increase until inflation is seen to have peaked. Because of
the perceived market distress, the mortgage rates and Treasury notes
will continue to become more positively correlated until the market can
show it is not distressed (inflation will need to show it has peaked for
this to happen). This sentiment is reflected in National Association of
Realtors Chief Economist Lawrence Yun's comments when he said "Today's
announcement by the Federal Reserve set a big increase in interest rates
and means several more rounds of rate hikes are on the way in upcoming
months ... rental demand will strengthen along with rents. Only when
consumer price inflation tops out and starts to fall will mortgage rates
stabilize or even decline a bit."

The expectation after the recent Fed Rate hike and the underwhelming
housing starts/permit number report this past week seems to be that the
mortgage rates will keep climbing until inflation is shown to have
reached a peak. Currently, the mortgage bonds and US Treasury Bonds are
exhibiting an increasing positive correlation that shows that the
mortgage market's performance is dependent upon inflation subsiding
before mortgage rates can stabilize.

Resources

<https://www.moodysanalytics.com/-/media/whitepaper/2014/2014-29-08-PCA-for-Yield-Curve-Modelling.pdf>

<https://insightr.wordpress.com/2017/04/14/american-bond-yields-and-principal-component-analysis/>

<https://arxiv.org/ftp/arxiv/papers/1911/1911.07288.pdf>
<https://www.r-bloggers.com/2021/12/easy-interpretations-of-adf-test-in-r/>

<http://nakisa.org/bankr-useful-financial-r-snippets/principal-component-analysis/>

<https://research-doc.credit-suisse.com/docView?language=ENG&format=PDF&source_id=csplusresearchcp&document_id=1001969281&serialid=EVplkK6oNi2Oum067aSBs%2Bp%2F04%2F3pgbDBc%2B1pGHrQ0U%3D&cspId=null>

<http://www.bondeconomics.com/2018/12/primer-understanding-principal.html>

<https://online.stat.psu.edu/stat505/lesson/11/11.4>
